---
title: "262 HW 3"
author: "Molly Williams"
date: "5/24/2019"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background

Goal: Compare policy options by analyzing potential average power output if rolling coefficients are reduced by 50% OR mean speed is reduced from 100 km/h to 80 km/h. 


### Mean power calculation for 100 km/h (27.8 m/s) and 50% reduction in c_roll
```{r}

# Source the R file that contains function auto() and includes all of the pre-set parameters for calculating vehicle power
source("my_function.R")

# Establish probabilities for car areas and mass
possible_cars = data.frame(mass=c(31000,45000,38000), area = c(25,30,22))
possible_cars$prob = c(0.4, 0.4, 0.2)

# Create array of 100 samples from a distribution with a mean of 27.8 m/s and 10% standard deviation (assuming log normal distribution of speeds around this mean), store in a new df called results
nsample = 100000
set.seed(262)
m_100 = 27.8
speeds_100 = rnorm(mean=m_100, sd=0.1*m_100, n=nsample)
results_100 = data.frame(speed=speeds_100, power=NA)


# Assign mass and area to each car based on probability, add into results df 
possible_cars$row = seq(from=1, to=nrow(possible_cars))
whichcar = base::sample(possible_cars$row, size=nsample, prob=possible_cars$prob, replace=TRUE)
results_100$mass = possible_cars$mass[whichcar]
results_100$area = possible_cars$area[whichcar]

# Calculate power (W) for each sample in the distribution using rolling coefficient that is reduced by 50% from 0.015 to 0.0075
results_100$power = auto(A=results_100$area, V=results_100$speed, m=results_100$mass, c_roll = 0.0075)

# Find mean of the power estimates (W)
print(mean(results_100$power))

# 182,679 W = average power output

```



### Mean power calculation for 80 km/h (22.2 m/s)
```{r}


# Create array of 100 samples from a distribution with a mean of 22.2 m/s and 10% standard deviation (assuming log normal distribution of speeds around this mean), store in a new df called results
nsample = 100000
set.seed(262)
m_80 = 22.2
speeds_80 = rnorm(mean=m_80, sd=0.1*m_80, n=nsample)
results_80 = data.frame(speed=speeds_80, power=NA)


# Add car mass and area based on probability into new results df 
possible_cars$row = seq(from=1, to=nrow(possible_cars))
whichcar = base::sample(possible_cars$row, size=nsample, prob=possible_cars$prob, replace=TRUE)
results_80$mass = possible_cars$mass[whichcar]
results_80$area = possible_cars$area[whichcar]

# Calculate power for each sample in the distribution using original rolling coefficient of 0.015
results_80$power = auto(A=results_80$area, V=results_80$speed, m=results_80$mass)

# Find mean of the power estimates 
print(mean(results_80$power))

# 177,447 W = average power output

```

Reducing average speed results in a lower power output, and therefore lower fuel consumption, making this the more favorable policy option.

