---
title: "Multidimensional_arrays"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Multidimensional Arrays


A vector is a 1 dimensional array
  * c(1,2,3,9,10)
A matrix is a 2 dimensional array

But what if we want  many arrays?

lets say an array is uses rows for locations, columns for animals, and each cell records the number of each
animal at each location
What if we want to look at sample from different years - we need another "dimension"

location * animal * time 
(3 dimensions)

what if samples were done by different researchers (4 dimensions)

location * animal * time * research

One way to represent data that is organized by different categories (animal, time, location etc) is to use
multi-dimensional arrays

Alternatively you can use a single data frame, but there are advantages to arrays - so worth learning...
* the single data frame will need a column for each dimension

#Creating the arary

Lets imagine we are sampling soil moisture on differrnt plots types (avocado, apple, orange) on different farms
and soil moisture is sampled at two different depths (25cm and 0.5 m)

Thats 3 dimension
*plot type (4) 
*farm (5)
*depth (2)




```{r creating}
# soil moisture at points for 5 different farms
# and 4 different plots within each farm, 
# one from avocado, one from apple, one from orange
# one from almond - so 4 fruits
# store soil moisture at 2 depths 
# ( 25cm, 0.5m)

# create array
soilm = array(dim=c(5,4,2))
dim(soilm)
soilm


# populate with values (we are just guessing, ideally this is where measuring would occur)
for (i in 1:5) {
  for (j in 1:4) {
    value = runif(min=0.2,max=0.5,n=2)
    soilm[i,j,]=value
  }
}
soilm

# add useful names
dimnames(soilm) = list(c("Farm1","Farm2","Farm3","Farm4","Farm5"), 
                      c("apple","avocado","orange","almond"),
                      c("shallow","deep"))



```

## Using multi-dimensional arrays

Ok so now lets look at the data stored in the arrays in different ways

Use *apply* - and think about which dimensions to **keep** in the apply

```{r using, echo=TRUE}

# overall whole area and all depths
# easiest!
averagesoilm = mean(soilm)
averagesoilm

#lets say we want to keep the locations (farms and plots) but average 
# over the two depths - average over the 3rd dimension
averagegrid = apply(soilm, c(1,2),mean)
averagegrid

# now lets say we want to also find the average for each farm
# keep the first dimension (5-farms) but average over plots and depths
averagefarm = apply(soilm, c(1), mean)
averagefarm

# average by fruit
averagefruit = apply(soilm, c(2), mean)
averagefruit

# average by  soil moisture depth
averagedepth = apply(soilm, c(3), mean)
averagedepth

# you can use other functions as well
sumfruit = apply(soilm, c(2), sum)
sumfruit

#  use your own function to summarize data 
isdrought = function(sm, thresh) {
  n = length(sm)
  drought=TRUE
  for (i in 1:n) {
    if (sm[i] > thresh) drought=FALSE
  }
  return(drought)
}
soilm
#local drought
apply(soilm, c(1,2), isdrought, thresh=0.4)

#farm drought
apply(soilm, c(1), isdrought, thresh=0.5)

#total area  is there drought anywhere
isdrought(soilm, thresh=0.3)
isdrought(soilm, thresh=0.7)
```


Finally we may find it easier to work with a simple data frame
We can always convert to that

```{r convert}

library(reshape2)
soilm_asdf = melt(soilm)
soilm_asdf
colnames(soilm_asdf)=c("Farm","Fruit","Depth","sm")
ggplot(soilm_asdf, aes(Farm, sm, fill=Fruit))+geom_col(position="dodge")+facet_wrap(~Depth)



```

